---
title: "Analysis of the Health and Economic Consequences of Severe Weather Events"
author: "AmolSahasrabudhe"
date: "March 5, 2017"
output: html_document
---

```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
This document describes an the data, transformations and analysis that allows us to evaluate the economic and health impacts of extreme weather events such as storms. The data was made availabe from the NOAA website

### Output session info &  Include necessary packages upfront
```{r message=FALSE}
sessionInfo()
library(dplyr)
library(ggplot2)
```

## Data Processing
```{r cache=TRUE}
stormfilelink = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(stormfilelink, destfile = "./StormData.bz2")
stormData<- tbl_df(read.csv(bzfile("./StormData.bz2")))
```
The data set has `r ncol(stormData)` variables and `r nrow(stormData)` observations and there are 985 types of Events grouped by the EVTYPE variable. 

## Analysis
Organize and group the data by the type of consequence fatalities, injuries
```{r}
groupByEvents <- group_by(stormData, EVTYPE)
aggResults <- data.frame(summarise(groupByEvents, 
                        totalfatalities = sum(FATALITIES), 
                        totalinjuries = sum(INJURIES), 
                        totalpropertydamages = sum(PROPDMG)))
top10fatalities <- head(arrange(aggResults, desc(totalfatalities)), 10)
top10injuries <- head(arrange(aggResults, desc(totalinjuries)), 10)
top10property <- head(arrange(aggResults, desc(totalpropertydamages)), 10)

```

## Charts

### first plot totalfatalities by Event type
```{r}
g <- ggplot(data=top10fatalities, aes(x=EVTYPE, y=totalfatalities))
g + geom_bar(stat="identity" , aes(fill=totalfatalities)) +
  scale_fill_gradient(low="cyan",high="red") +    theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
```
### Plot totalInjuries by Event type
```{r}
g <- ggplot(data=top10injuries, aes(x=EVTYPE, y=totalinjuries))
g + geom_bar(stat="identity" , aes(fill=totalinjuries)) +
  scale_fill_gradient(low="yellow",high="orange") +    theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
```

### Plot totalPropertyDamages by Event type
```{r}
g <- ggplot(data=top10property, aes(x=EVTYPE, y=totalpropertydamages))
g + geom_bar(stat="identity" , aes(fill=totalpropertydamages)) +
  scale_fill_gradient(low="light blue",high="dark blue") +    theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
```
## Results
The charts above would suggest that Tornadoes are the number one reason for fatalities and injuries. Looking at the property damages the conclusion is the same. Tornadoes lead to the most property damage in totality across the country.
